---
title: "EXAMINATION PROJECT"
author: "Dinh_Long Huynh"
date: "Nov 27, 2023"
email: "Dinh-Long.Huynh.4289@student.uu.se"
version: 2
output: 
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    code_folding: show
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smoot = tidyh_scroll: yes
mainfont: null
sansfont: null
mathfont: null
monofont: null
fontsize: null
---

# TASK 1: IMPORT DATA FILES

## Load library

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(GGally)
library(knitr)
library(PKNCA)
```

## Import data and constant

```{r}
# Set flexible directory links
dir_name <- paste0(getwd(), "/Original_data/")

pk_name <- "BPI889_PK_29.csv"
pk_path <- paste0(dir_name, pk_name)

snp_name <- "BPI889_SNP_29.txt"
snp_path <- paste0(dir_name, snp_name)


#Import pk data. In pk data, there are several missing value, represented as ".", due to the missing values in real study. Thus, the argument na = "." is required
data_pk <- read.csv(pk_path, header = T, na = ".")
str(data_pk)
kable(head(data_pk, 10), 
      align = "c", 
      caption = "data_pk")

#Import snp data. The row.names = NULL argument is used for indicating that the 1st columns is belong to data frame, not the row names column.
#The read.table() is used because it can automatically recognize the separation between column without sep = "". 
data_snp <- read.table(snp_path, header = T, row.names = NULL)
str(data_snp)
kable(head(data_snp, 10), 
      align = "c", 
      caption = "Part of data_snp")

# Import dose as the constant.
dose <- 200
```

# TASK 2: CHANGE VARIABLES NAMES, DATA FORMAT, AND DATA TYPES
## Manipulating data_pk
```{r}
# Change 1st column of data_pk into ID
colnames(data_pk)[1] <- "ID"

# Change SEX of data_pk column into factor.
data_pk$Sex <- factor(data_pk$Sex, 
                      levels = c("M", "F"), 
                      labels = c("Male", "Female"))

# Change NA value at Time 0.15 = 0 to avoid error when calculate anoter PK variables
data_pk$Time.0.15.h = 0

#Because the data_pk is recently the wide format, so it is better to change into long format before merging.
tidy_pk <- gather(data_pk, 
                  key = "Time", 
                  value = "Concentration", 
                  -c("ID","Sex", "Weight..kg.", "Height..cm.", "Age..yrs."))

#Time is continous variables, so it is better to change data type of Time into numeric.
tidy_pk$Time <- gsub("Time.", "", tidy_pk$Time)
tidy_pk$Time <- gsub(".h", "", tidy_pk$Time)
tidy_pk$Time <- as.numeric(tidy_pk$Time)

kable(head(tidy_pk, 10), 
      align = "c", 
      caption = "Part of tidy_pk: the long format of data_pk")
```

## Manipulating data_snp
```{r}
#Because the data_snp is recently the wide format, so it better to change into long format before merge.
colnames(data_snp)[1] <- "ID"
tidy_snp <- gather(data_snp, key = "SNP", value = "Type", -ID)
tidy_snp$Type <- factor(tidy_snp$Type, 
                        levels = c(0, 1, 2), 
                        labels = c("Wildtype", "Heterogeneous", "Homogeneous"))
kable(head(tidy_snp, n = 10), 
      align = "c", 
      caption = "Part of tidy_snp: the long format of data_snp")
```

# TASK 3: COMBINE PK AND SNP

```{r}
data_all <- merge(tidy_pk, tidy_snp, by = "ID")

#Reorder the columns
data_all <- select(data_all, 
                   ID, Sex, Weight..kg., Height..cm., Age..yrs.,SNP, Type, Time, Concentration) 

#Reorder the rows
data_all$ID <- as.numeric(gsub("pat", "", data_all$ID))
data_all <- arrange(data_all, ID,Time)
data_all$ID <- factor(data_all$ID, 
                      levels = c(1:100))

#The samples data frame: 
kable(head(data_all, 10), 
      align = "c", 
      caption = "data_all")
```

# TASK 4: CALCULATE BODY COMPOSITION MEASEURMENT: TBW

```{r}
data_all$TBW <- NA
data_all$TBW[data_all$Sex == "Male"] <- round(2.447-0.09156*data_all$Age..yrs. + 0.1074*data_all$Height..cm. + 0.3362*data_all$Weight..kg., 1)
data_all$TBW[data_all$Sex == "Female"] <- round(-2.097 + 0.1067*data_all$Height..cm. + 0.2466*data_all$Weight..kg.,1)
```

# TASK 5: CATEGORIZE TBW INTO TWO GROUPS

```{r}
data_all$CTBW <- NA
data_all$CTBW[data_all$TBW > 40] <- "above 40L"
data_all$CTBW[data_all$TBW < 40] <- "below 40L"

#The samples data frame: 
kable(head(data_all, 10), 
      align = "c", 
      caption = "Part of data_all after adding TBW and CTBW")
```

# TASK 6: CALCULATE CMAX, T1/2, AND CL

## Method 1: Linear regression in the linear range of time

### Find k and Cmax

```{r}
#Change format of ID column in tidy_pk into the same format with data_all
tidy_pk$ID <- as.numeric(gsub("pat", "", tidy_pk$ID))
tidy_pk <- arrange(tidy_pk, ID,Time)
tidy_pk$ID <- factor(tidy_pk$ID, 
                     levels = c(1:100))

#Determine ln(C) values
tidy_pk$log_con <- log(tidy_pk$Concentration)

#The pharmacokinetic parameters are specified for each ID, thus, the group_by function is used.
#Filter function is used for chosing the time after tmax (approximately 1 hour) and avoid -Inf values
grouped_by_ID <- tidy_pk%>% filter(Time>1 & Time < 12 )%>% group_by(ID) %>%
  summarise(k = lm(log_con ~ Time)$coefficients[2], #linear regression and extract the slope (k)
            Co = exp(lm(log_con ~ Time)$coefficients[1])) #linear regression and extract the intercept, which is ln(Co), and then use exp(ln(Co)) to calculate Co
            
#Cmax determination
grouped_by_ID_Cmax <- tidy_pk%>%group_by(ID)%>%
  summarise(Cmax = max(Concentration, na.rm = T))

#At this point, I have two differents data frame of several grouped parameters, then it should be merge together
grouped_by_ID <- merge(grouped_by_ID,grouped_by_ID_Cmax, by = "ID")
grouped_by_ID <- arrange(grouped_by_ID, ID)
```

### Find t1/2 and CL

```{r}
grouped_by_ID$half.life <- log(2)/abs(grouped_by_ID$k)
grouped_by_ID$Vd <- dose/grouped_by_ID$Co
grouped_by_ID$CL <- abs(grouped_by_ID$k)*grouped_by_ID$Vd

#The samples data frame: 
kable(head(grouped_by_ID, 10), 
      align = "c",
      digits = 3,
      caption = "Part of grouped_by_ID after calculating pharmacokinetic parameters")
```

## Method 2: Use PKNCA package
```{r}
#Create the Dose column in tidy pk and extract the dose data frame corresponding to the starting time.
tidy_pk$Dose <- 200
dose_data <- tidy_pk %>% select(ID, Time, Dose) %>% filter(Time == 0.15)

#To run PKNCA calculator, it is neccessary to create PKNCA_dose and PKNCA_concentration data frame first.
PKNCA_dose <- PKNCAdose(dose_data, Dose~Time | ID)
PKNCA_concentration <- PKNCAconc(tidy_pk, Concentration~Time|ID)

#Compete data preparation
PKNCA_data <- PKNCAdata(PKNCA_concentration,PKNCA_dose)

#Load calculation
PKNCA_results <- pk.nca(PKNCA_data)

#Results visualization
Raw_PK_variables <- PKNCA_results$result
PK_variables <- Raw_PK_variables %>% group_by(ID) %>% 
  summarise(k = -log(2)/PPORRES[PPTESTCD == "half.life"],
            Cmax = PPORRES[PPTESTCD == "cmax"],
            half.life = PPORRES[PPTESTCD == "half.life"],
            CL = dose/PPORRES[PPTESTCD == "auclast"])

#The samples data frame: 
kable(head(PK_variables, 10), 
      align = "c", 
      digits = 3,
      caption = "Part of PK_variables")
```

# TASK 7: SUMMARIZE CMAX, T1/2, AND CL
## Summary from linear regression method
```{r}
summary_parameter_lm = data.frame()
#Loop over parameters, determine summary variables, and at the same time, store them into data frame.
for (parameter in c("half.life", "Cmax", "CL")) {
  summary_parameter_lm["Mean", parameter] = mean(grouped_by_ID[[parameter]])
  summary_parameter_lm["Standard Deviation", parameter] = sd(grouped_by_ID[[parameter]])
  summary_parameter_lm["Median", parameter] = median(grouped_by_ID[[parameter]])
  summary_parameter_lm["Max", parameter] = max(grouped_by_ID[[parameter]])
  summary_parameter_lm["Min", parameter] = min(grouped_by_ID[[parameter]])
}

#Create and print summary data frame
kable(summary_parameter_lm, 
      align = "c", 
      digits = 3,
      caption = "Summary of half life, Cmax, and CL")

```
## Summary from PKNCA packages method
```{r}
summary_parameter_PKNCA = data.frame()
#Loop over parameters, determine summary variables, and at the same time, store them into data frame.
for (parameter in c("half.life", "Cmax", "CL")) {
  summary_parameter_PKNCA["Mean", parameter] = mean(PK_variables[[parameter]])
  summary_parameter_PKNCA["Standard Deviation", parameter] = sd(PK_variables[[parameter]])
  summary_parameter_PKNCA["Median", parameter] = median(PK_variables[[parameter]])
  summary_parameter_PKNCA["Max", parameter] = max(PK_variables[[parameter]])
  summary_parameter_PKNCA["Min", parameter] = min(PK_variables[[parameter]])
}

#Create and print summary data frame
kable(summary_parameter_PKNCA, 
      align = "c", 
      digits = 3,
      caption = "Summary of half life, Cmax, and CL")

```

## Compare two methods
```{r}
t.test(grouped_by_ID$half.life, PK_variables$half.life)
```
p-value of t-test for half life is greater than 0.05, indicating for no significant difference in half life between two methods. 

```{r}
t.test(grouped_by_ID$Cmax, PK_variables$Cmax)
```
p-value of t-test for Cmax is 1, however, the Cmax values are exactly the same between two methods. Therefore, there is no difference in Cmax between two methods. 

```{r}
t.test(grouped_by_ID$CL, PK_variables$CL)
```
p-value of t-test for CL is smaller 0.05, indiciating for the significant difference in CL between two methods. Indeed, two methods used two different algorithm to calculate CL. 

For linear regression, CL is calculated from Dose, Co and k. 
For PKNCA package, CL is calculated from Dose and AUC. 

However, the Co of oral drug is just the theoretical concentration assumped when the absorption occurs quickly (similar to intravenous drug), derrived from the intercept of linear regression. Thus, using Dose and AUC, which means PKNCA package to calculate PK variables is more accurate.  


# TASK 8: GRAPHICALLY DISPLAY INDIVIDUAL CONCENTRATIONS OF BPI889 VERSUS TIME (SPAGHETTI PLOT)

```{r}
ggplot(tidy_pk, aes(x=Time, y=Concentration, group = ID, color = ID))+
  geom_line(na.rm=T)+
  xlab("Time (hours)")+
  ylab("Concentration (mg/L)")+
  ggtitle("Concentration versus Time for each ID")+
  theme_classic()

ggplot(tidy_pk, aes(x=Time, y=log_con, group = ID, color = ID))+
  geom_line(na.rm=T)+
  xlab("Time (hours)")+
  ylab("ln(Concentration)")+
  ggtitle("ln(Concentration) versus Time for each ID")+
  theme_classic()
```

# TASK 9: GRAPHICALLY DISPLAY CORRELATIONS BETWEEN CMAX, T1/2 AND CL (SCATTER PLOT)

## Basic way

```{r}
data_all <- merge(data_all, PK_variables, by = "ID")
pairs(data_all[,c("Cmax", "half.life", "CL")], 
      upper.panel = NULL, 
      lower.panel = panel.smooth)
```

## Ggplot way

```{r}
ggpairs(data_all[,c("Cmax", "half.life", "CL")])
```

# TASK 10: GRAPHICALLY DISPLAY T1/2 AND CL VERSUS SNPS (BOX-WHISKERS PLOTS)

## Basic way

```{r}
par(mfrow=c(2,5))
color <- c("red4", "green", "blue")
for (SNP in unique(data_all$SNP)) {
  boxplot(half.life~factor(Type,levels = c("Wildtype", "Heterogeneous", "Homogeneous"), 
                           labels = c("Wi", "He", "Ho")), 
          data_all[data_all$SNP == SNP, ],
          ylab = "Half life (hours)",
          xlab = "Mutations",
          col = color,
          main = SNP)
}

for (SNP in unique(data_all$SNP)) {
  boxplot(CL~factor(Type,levels = c("Wildtype", "Heterogeneous", "Homogeneous"), 
                           labels = c("Wi", "He", "Ho")), 
          data_all[data_all$SNP == SNP, ],
          ylab = "Clearance (L/h)",
          xlab = "Mutations",
          col = color,
          main = SNP)
}



```

## Ggplot way

```{r}
half_life_boxplot <- ggplot(data_all, aes(x = factor(Type, 
                                                     levels = c("Wildtype", "Heterogeneous", "Homogeneous"), 
                                                     labels = c("Wi", "He", "Ho")), 
                                          y = half.life, fill = Type))+
  geom_boxplot()+
  ylab("Half life (hours)")+
  xlab("Mutations")+ 
  ggtitle("Box plot of Half-life versus SNPs ")+
  facet_grid(~SNP)+
  theme_classic()

clearance_boxplot <- ggplot(data_all, aes(x = factor(Type, 
                                                     levels = c("Wildtype", "Heterogeneous", "Homogeneous"), 
                                                     labels = c("Wi", "He", "Ho")), 
                                          y = CL, fill = Type))+
  geom_boxplot()+
  theme_classic()+
  ylab("Clearance (L/h)")+
  xlab("Mutations")+ 
  ggtitle("Box plot of Clearance versus SNPs ")+
  facet_grid(~SNP)

ggarrange(half_life_boxplot, clearance_boxplot, nrow = 2, ncol =1)
```

# TASK 11: GRAPHICALLY DISPLAY CORRELATIONS BETWEEN T1/2 AND TBW TO ASSESS A RELATIONSHIP AND ADD A LINEAR REGRESSION (SCATTER WITH LINEAR REGRESSION)

## Basic way

```{r}
plot(x = data_all$TBW, y = data_all$half.life,
     xlab = "Total body water (L)",
     ylab = "Half life (hours)",
     main = "Correlation between TBW and Half life",
     pch = 16)

TBW_halflife_regression = lm(half.life~TBW, data_all)
abline(coef(TBW_halflife_regression), lwd = 2, col = "blue")
```

## Ggplot way

```{r}
ggplot(data_all, aes(x=TBW, y = half.life))+ 
  geom_point()+
  geom_smooth(method = "lm", na.rm = T, se = F)+
  xlab("Total body water (L)")+
  ylab("Half life (hours)")+
  theme_classic()+
  ggtitle("Correlation between TBW and Half life")
```

# TASK 12: PERFORM AN ANOVA OF CL AND CMAX FOR THE FIVE SNPS

## ANOVA for CL between SNPs

```{r}
CL_SNP_aov <- aov(CL~SNP, data_all)
summary(CL_SNP_aov)
```

p-value = 1. Conclusion: There is not enough evidence to reject the "no significant difference" in clearance between SNPs.

## ANOVA for Cmax between SNPs

```{r}
Cmax_SNP_aov <- aov(Cmax~SNP, data_all)
summary(Cmax_SNP_aov)
```

p-value = 1. Conclusion: There is not enough evidence to reject the "no significant difference" in clearance between SNPs.

# TASK 13: PERFORM A T-TEST OF T1/2 FOR THE TWO CATEGORICAL GROUPS OF TBW

```{r}
t.test(half.life~CTBW, data_all)
```

p-value > 0.05. Conclusion: There is no significant difference in drug half life between the groups having TBW above 40L and the groups having TBW below 40L.
