---
title: 'Practice Project'
subtitle: 'R session'
author: 'Dinh Long Huynh'
date: 'Nov 16, 2023'
output: 
  html_document: 
    code_folding: show
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Setup environment and data import
## Load libraries
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(GGally)
```

## Set working directory
```{r}
dir_name <- paste0(getwd(),"/Original_data/")
```

## Import the datasets
```{r}
data_pd <- read.csv(paste0(dir_name, "BPI889_PD_run-in.csv"), header = TRUE, dec = ".", na = ".", as.is = TRUE)
data_cov <- read.table(paste0(dir_name, "BPI889_demographics.tab"), header = TRUE, dec = ".", na = ".", as.is = TRUE)
```

## Visually inspect imported data.frames
```{r}
str(data_pd)
head(data_pd, n=10)
str(data_cov)
head(data_cov, n=10)
```

# Data Management
## Change variables
### Set variables names
```{r}
colnames(data_pd)[1] <- "ID"
```

### Transform dataset from wide to long format
```{r}
tidy_pd <- gather(data_pd, key = TIME, value = CD4, -ID)
print(paste("tidy_pd has", nrow(tidy_pd),"rows and",ncol(tidy_pd),"columns"))
```

### Format Numeric variables
```{r}
tidy_pd$TIME <- gsub(pattern = 'day.', replacement = '', x = tidy_pd$TIME, ignore.case = T)
unique(tidy_pd$TIME)
tidy_pd$TIME <- as.numeric(tidy_pd$TIME)
str(tidy_pd)
print("The TIME has been already numeric data. However, it should be discrete data for further steps, thus I changed it into factors")
tidy_pd$TIME <- as.factor(tidy_pd$TIME)
str(tidy_pd)
```

### Formate categorical variables
```{r}
data_cov$SEX <- factor(data_cov$SEX, levels = c("M", "F"), labels = c("Male", "Female"))
data_cov$TB <- factor(data_cov$TB, levels = c(0, 1), labels = c("HIV", "HIV+TB"))
str(data_cov)
print("The SEX and TB have been already factor data.")
```

### Create new variables
```{r}
summary(data_cov$HT)
print("Given the range of HT, the column have unit of cm")
summary(data_cov$WT)
print("Given the range of WT, the column have unit of kg")

data_cov <- mutate(data_cov, BMI = round(WT/((HT/100)^2),1))
data_cov$CBMI = NA
data_cov$CBMI[data_cov$SEX == "Male" & data_cov$BMI == 20] <- 0
data_cov$CBMI[data_cov$SEX == "Female" & data_cov$BMI == 18] <- 0
data_cov$CBMI[data_cov$SEX == "Male" & data_cov$BMI > 20 & data_cov$BMI <= 25] <- 1
data_cov$CBMI[data_cov$SEX == "Female" & data_cov$BMI > 18 & data_cov$BMI <= 25] <- 1
data_cov$CBMI[data_cov$BMI > 25 & data_cov$BMI <= 30] <- 2
data_cov$CBMI[data_cov$BMI > 30] <- 3
data_cov$CBMI <- factor(data_cov$CBMI, levels = c(0,1,2,3), labels = c("underweight", "lean", "overweight","obese" ))
table(data_cov$CBMI)
```

## Merge datasets
```{r}
data_cbind <- cbind(tidy_pd, data_cov)
data_merge <- merge(tidy_pd, data_cov, all = TRUE)
data_all <- merge(tidy_pd, data_cov, by ="ID")
summary(data_cbind)
summary(data_merge)
summary(data_all)
```

## Re-arrange the data
### Reorder variables
```{r}
data_all <- select(data_all, ID, TIME, CD4, BCD4, BVL, WT, HT, BMI, CBMI, SEX, TB)
colnames(data_all)
```

### Reorder rows
```{r}
data_all <- arrange(data_all, ID, TIME)
```

# Numerical and graphical data summary
## Numerical summary of time-invariant variables
```{r}
data_first <- data_all[!duplicated(data_all$ID), ]
summary(data_first[,c("BCD4", "BVL", "WT", "HT","BMI")])
summary(data_first[,c("CBMI", "SEX","TB")])
```

## Graphical summary of time-invariant variables
### Histograms
#### Basic plot
```{r}
par(mfrow = c(3, 2))
for(i in c("BCD4", "BVL","WT", "HT", "BMI")) {
  hist(data_first[[i]], 
       main = paste("Histogram of",i),
       xlab = i)
}
```

#### ggplot 
```{r}
histogram_list <- list()
for(i in c("BCD4", "BVL","WT", "HT", "BMI")) {
  histogram_list[[i]] = ggplot(data_first, aes_string(i))+
           geom_histogram(fill="gray", color="black")+
           ggtitle(paste("Histogram of",i))+
           xlab(i)+
          theme_classic()
}
ggarrange(plotlist = histogram_list, 
          ncol = 2, nrow = 3)
# Keep in mind: aes() works with column, not “column”. 
#Alternative: aes_string() to extract only the string inside “ “
#or .data(“column”)
```

### Correlation matrix
#### Basic plot
```{r}
pairs(data_first[,c(4:8)],
      upper.panel = NULL,
      lower.panel = panel.smooth)
```

#### ggplot 
```{r}
ggpairs(data_first[,c(4:8)])+
  theme_classic()
```

### Box plots
#### Basic plot
```{r}
colours <- c('dodgerblue1','skyblue1','skyblue2','skyblue3')
boxplot(WT ~ CBMI,
        data = data_first,
        main = 'Box plot of WT by CBMI',
        col = colours,
        notch = T,
        cex = 0.8)
legend(x = 0.5, y = 130,
       legend = levels(data_first$CBMI), bty = "n",
       col = "black", lwd = 0.5, pch = 22, cex = 0.8, fill = colours)
```

#### ggplot
```{r}
ggplot(data_first, aes(x=TB, y=BCD4, fill = TB))+
  geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2, notch = TRUE)+
  ggtitle("Box plot of BCD4 by TB")+
  scale_fill_brewer(palette = "Blues") +
  theme_classic()
```

## Numerical summary of time-variant variables
```{r}
grouped_data_all <- data_all%>% group_by(TIME) %>% summarise('mean_value' = mean(CD4, na.rm = T), 'median_value' = median(CD4, na.rm = T), 'sd_value' = sd(CD4, na.rm = T), 'Min' = min(CD4, na.rm=T), 'Max' = max(CD4, na.rm = T))
grouped_data_all
```

## Graphical summary of time-variant variables
### Scatterplots
#### Spaghetti plots
```{r}
ggplot(data_all, aes(x=TIME, y=CD4, color = TB, group = ID))+
  geom_line(na.rm=T)+
  coord_cartesian(ylim=c(100, 500))+
  geom_hline(yintercept = 500, color = "red")+
  geom_hline(yintercept = 200, color = "red")
```

#### Panel plots
```{r}
ggplot(data_all, aes(x=TIME, y=CD4, color = TB))+
  geom_point( na.rm = T)+
  facet_wrap(~SEX)
```

# Statistical testing
```{r}
CD4_annova <- aov(CD4~TIME, data = data_all)
summary(CD4_annova)
print("p-value= 0.981 >> 0.05. Conclusion: There is no significant difference of CD4 count on day 0, 7, and 14")

ttest_data <- data_all%>%filter(TIME == "0")%>% select(CD4, TB)
t.test(CD4~TB, data = ttest_data)
print("p-value= 9.945e-06 << 0.05. Conclusion: There is significant difference of CD4 count between HIV group and HIV+TB group")
```

